\Rpackage{MSnbase}: labelled MS2 data pre-processing, visualisation and quantification.

### 2012-02-25 inst/doc/MSnbase-demo/MSnbase-demo.Rnw
more MSnSet combine, new updateFeatureNames
 o new updateFeatureNames function <2012-02-17 Fri>
 o updated vignettes to illustrate vertical/horizontal combine <2012-02-17 Fri>
 o typo in normalised.Rd <2012-02-19 Sun>
 o TODO combine unit tests




git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MSnbase@63146 bc3139a8-67e5-0310-9ffc-ced21a209358

Default MSnSet feature names (\Robject{X1}, \Robject{X2}, \ldots) and sample names 
(\Robject{iTRAQ4.114}, \Robject{iTRAQ4.115}, \Robject{iTRAQ4.116}, \ldots) are not very informative.
The features and samples of these anonymous quantitative data-sets should be updated prior 
to additional downstream analysis.

Lets create two \Robject{MSnSet}s from the same raw data to simulate two different 
independent experiments. As done previously (see section \ref{sec:feataggregation}), 
we combine the specta based on the proteins they have been identified to belong to.
Features can thus naturally be named using protein accession numbers.
Alternatively, if peptide sequences would have been used as grouping factor in 
\Rfunction{combineFeatures}, then these would be good feature name candidates.

<<make2exps, echo=TRUE, cache=FALSE>>=
set.seed(1)
i <- sample(length(itraqdata), 35)
j <- sample(length(itraqdata), 35)
exp1 <- quantify(itraqdata[i], reporters = iTRAQ4, verbose = FALSE)
exp2 <- quantify(itraqdata[j], reporters = iTRAQ4, verbose = FALSE)
table(featureNames(exp1) %in% featureNames(exp2))
exp1 <- combineFeatures(exp1, groupBy = fData(exp1)$ProteinAccession)
exp2 <- combineFeatures(exp2, groupBy = fData(exp2)$ProteinAccession)
head(featureNames(exp1))
featureNames(exp1) <- fData(exp1)$ProteinAccession
featureNames(exp2) <- fData(exp2)$ProteinAccession
head(featureNames(exp1))
@

@@

Sample names should also be updated to replace reporter names with relevant identifiers; 
the individual reporter data is stored in the \Robject{phenoData} and is not lost.
A convenience function \Rfunction{updateSampleNames} is provided to append the \Robject{MSnSet}'s 
variable name to the already defined names, although in general, biologically relevant 
identifiers are preferred.

@@

At this stage, it is not yet possible to combine the two experiments, 
because their feature data is not compatible; they share the same feature 
variable labels, i.e. the feature data column names 
(\Sexpr{paste(head(fvarLabels(exp1), n = 3), collapse=", ")}, \ldots), 
but the content is different because the original data was, in particular 
all the spectrum data. Feature data with identical labels (columns in the data frame) 
and names (row in the data frame) are expected to have the same data and produce an
error if not conform.

@@

Instead of removing these overlapping and different feature data columns, 
one can use a second convenience function, \Rfunction{updateFvarLabels}, 
to update feature labels based on the experiements variable name and maintain all the metadata.

@@

It is now possible to combine \Robject{exp1} and \Robject{exp2} with the 
\Rfunction{combine} method. The new experiment will contain the union of 
the feature names of the individual experiments with missing values 
inserted appropriately.

@@

In summary, when experiments need to be combined, one needs to 
(1) clarify the sample names using \Rfunction{updateSampleNames} or,
better manually, for biological relevance and (2) update the feature 
data variable labels with \Rfunction{updateFvarLabels}. 
The individual experiments (there can be more than 2) can then easily be combined 
with the \Rfunction{combine} method while retaining the meta-data.

>>>>

Combining mass spectrometry runs can be done in two different ways depending on the 
nature of these runs. If the runs represent repeated measures of identical samples, 
for instance multiple fractions, the data has to be combine along the row of the 
quantitation matrix: all the features (along the rows) represent measurements of the 
same set of samples (along the columns). In this situation, described in section 
\ref{sec:comb1}, two experiments of dimensions 
$n_1$ (rows) by $m$ (columns and $n_2$ by $m$ will produce 
a new experiment of dimensions $n_1 + n_2$ by $m$.

When however, different sets of samples have been analysed in different mass 
spectrometry runs, the data has 
to be combined along the columns of the quantitation matrix: some features will be 
shared across experiments and should thus be aligned on a same row in the new
data set, whereas unique 
features to one experiment should be set as missing in the other one. 
In this situation, described in section \ref{sec:comb2}, 
two experiments of dimensions $n_1$ by $m_1$ and $n_2$ by $m_2$ will 
produce a new experiment of dimensions 
$unique_{n_1} + unique_{n_2} + shared_{n_1, n_2}$ by $m_1 + m_2$. 
The two first terms of the first dimension will be complemented by \Robject{NA} values.

Default \Robject{MSnSet} feature names (\Robject{X1}, \Robject{X2}, \ldots) and sample names 
(\Robject{iTRAQ4.114}, \Robject{iTRAQ4.115}, \Robject{iTRAQ4.116}, \ldots) are not informative.
The features and samples of these anonymous quantitative data-sets should be updated 
before being combined, to guide how to meaningfully merge them.

\subsection{Combining identical samples}\label{sec:comb1}

To simulate this situation, let us use quantiation data from the \Robject{itraqdata} 
object that is provided with the package as experiment 1 and the data from the 
\Robject{rawdata} \Robject{MSnExp} instance created at the very beginning of this 
document. Both experiments share the \emph{same} default iTRAQ 4-plex reporter names 
as default sample names, and will thus automatically be combined along rows.

<<make2exps1, echo=TRUE, cache=FALSE>>=
exp1 <- quantify(itraqdata, reporters = iTRAQ4, verbose = FALSE)
sampleNames(exp1)
exp2 <- quantify(rawdata, reporters = iTRAQ4, verbose = FALSE)
sampleNames(exp2)
@

It important to note that the features of these independent experiments share 
the same default anonymous names: X1, X2, X3, \ldots, 
that however represent quantitation of distinct physical analytes. 
If the experiments were to be combined as is, it would result in an error because
data points for the same \emph{feature} name (say \Robject{X1}) and the 
same \emph{sample name} (say \Robject{iTRAQ4.114}) have different values. 
We thus first update the feature names to explicitate that they originate from 
different experiment and represent quantitation from different spectra 
using the convenience function \Rfunction{updateFeatureNames}.
Note that updating the names of one experiment would suffice here.

<<updateFnames, echo=TRUE, cache=FALSE>>=
head(featureNames(exp1))
exp1 <- updateFeatureNames(exp1)
head(featureNames(exp1))
head(featureNames(exp2))
exp2 <- updateFeatureNames(exp2)
head(featureNames(exp2))
@

The two experiments now share the same sample names and have different feature names 
and will be combined along the row. Note that all meta-data is correctly 
combined along the quantitation values.

<<comb1, echo=TRUE, cache=FALSE>>=
exp12 <- combine(exp1, exp2)
dim(exp1)
dim(exp2)
dim(exp12)
@

\subsection{Combine different samples}\label{sec:comb2}

Lets now create two \Robject{MSnSet}s from the same raw data to simulate two different 
independent experiments that share some features. 
As done previously (see section \ref{sec:feataggregation}), 
we combine the specta based on the proteins they have been identified to belong to.
Features can thus naturally be named using protein accession numbers.
Alternatively, if peptide sequences would have been used as grouping factor in 
\Rfunction{combineFeatures}, then these would be good feature name candidates.

<<make2exps2, echo=TRUE, cache=FALSE>>=
set.seed(1)
i <- sample(length(itraqdata), 35)
j <- sample(length(itraqdata), 35)
exp1 <- quantify(itraqdata[i], reporters = iTRAQ4, verbose = FALSE)
exp2 <- quantify(itraqdata[j], reporters = iTRAQ4, verbose = FALSE)
table(featureNames(exp1) %in% featureNames(exp2))
exp1 <- combineFeatures(exp1, groupBy = fData(exp1)$ProteinAccession)
exp2 <- combineFeatures(exp2, groupBy = fData(exp2)$ProteinAccession)
head(featureNames(exp1))
featureNames(exp1) <- fData(exp1)$ProteinAccession
featureNames(exp2) <- fData(exp2)$ProteinAccession
head(featureNames(exp1))
@

@@

Sample names should also be updated to replace anonymous reporter names with relevant identifiers; 
the individual reporter data is stored in the \Robject{phenoData} and is not lost.
A convenience function \Rfunction{updateSampleNames} is provided to append the \Robject{MSnSet}'s 
variable name to the already defined names, although in general, biologically relevant 
identifiers are preferred.

@@

At this stage, it is not yet possible to combine the two experiments, 
because their feature data is not compatible yet; they share the same feature 
variable labels, i.e. the feature data column names 
(\Sexpr{paste(head(fvarLabels(exp1), n = 3), collapse=", ")}, \ldots), 
but the part of the content is different because the original data was (in particular 
all the spectrum centric data: identical peptides in different runs will have different 
retention times, precursor intensities, \ldots). 
Feature data with identical labels (columns in the data frame) 
and names (row in the data frame) are expected to have the same data and produce an
error if not conform.

@@

Instead of removing these identical feature data columns, 
one can use a second convenience function, \Rfunction{updateFvarLabels}, 
to update feature labels based on the experiements variable name 
and maintain all the metadata.

@@

It is now possible to combine \Robject{exp1} and \Robject{exp2}, 
including all the meta-data, with the \Rfunction{combine} method. 
The new experiment will contain the union of 
the feature names of the individual experiments with missing values 
inserted appropriately.

@@

In summary, when experiments with different samples need to be combined 
(along the columns), one needs to 
(1) clarify the sample names using \Rfunction{updateSampleNames} or
better manually, for biological relevance and (2) update the feature 
data variable labels with \Rfunction{updateFvarLabels}. 
The individual experiments (there can be more than 2) can then easily be combined 
with the \Rfunction{combine} method while retaining the meta-data.

If runs for the same sample (different fractions for example) need to be combines, 
one needs to (1) differentiate the feature provenance with 
\Rfunction{updateFeatureNames} prior to use \Rfunction{combine}.
